---
title: "AXCPT Psychometrics"
output:  html_document
author: "Winnie Zhuang (wzhuang@live.ca)"
date: "11/17/20"
---

Some functions to assess the psychometrics of AXCPT's derived measures.

```{r}
library(tidyverse)
library(car)
library(kableExtra)
library(yarrr)
```

#Generate Data
First, let's make some data to work with.
```{r}
set.seed(1234)
dt <- expand.grid(
  ntrial=1:20, # number of trials/condition
  id= 1:80, #number of subjects
  trialtype= c("ax","ay","bx","by") #within conditions
)

#make 2 age groups as between subjects condition
dt$age_group<- NA
dt$age_group[dt$id<=40]<-1 
dt$age_group[dt$id>40]<-2

#simulate some rts and accs
dt$rt_ms<- rnorm(nrow(dt), mean=400, sd=200)
dt$acc<-sample(x=c(0,1),size=nrow(dt),replace=TRUE,prob=c(.2,.8))

#take a look at dt
dt %>% group_by(age_group,id,trialtype) %>%
  count()

#look at summary stats
dtage<-dt %>% group_by(age_group) %>% 
  summarise(subj_count= n_distinct(id),
            rt_mean= mean(rt_ms),
            rt_var=var(rt_ms),
            acc_mean=mean(acc),
            acc_var=var(acc)) 
dtage


```
In this simulated data, 2 age groups with 40 subjects each, 3 blocks of 20 trials per condition per block.

#Mean RTs
##Calculation
```{r}
#look at stats by id
dt_rt<-dt %>% group_by(id,trialtype,age_group) %>%
  summarise(rt_mean=mean(rt_ms))
```

##Variability
```{r}
#Levene's test to see if age groups and trialtypes have same variance
leveneTest(rt_ms~as.factor(age_group)*trialtype,
           data=dt,
           center=mean)

#make a plot of rt by trialtype and age group
pirateplot(formula= rt_mean~factor(age_group)*factor(trialtype),
           data=dt_rt,
           bean.f.o = .3, # customize bean (i.e. violin)
           bean.b.o= 0,
           point.o = 0.3,#customize points
            point.pch = 2,
           point.col = "black",
           point.cex = 0.1,
           inf.f.o = .3, # Inference fill
           #xlab="Age Group",#labels
           ylab="mean RT",
           theme=1)


```
No differences in variances between groups here. Makes sense since the data were generated from the same distribution. 

##Reliability
internal consistency reliability= how well the scores across instances/items are correlated
Do 

I used split-half reliabilities 
```{r rt split half}
#create a df with just the necessary info
dt_shr_rt<-dt %>%
  subset(select=c(id,age_group,trialtype,rt_ms)) #might also exclude incorrect trials. Here they've been included in the example. 

#now a df to store output
shr_meanrt<-data.frame(n_rep=integer(),
                    age_group=character(),
                    trialtype=character(),
                     meanrt_splithalf=double(),
                     meanrt_sb=double())

tot_rep<- 5000 #specify number of permutations per subject
cur_rep<-0

#now let's loop! This might take a while...
while (cur_rep<=tot_rep){
  cur_rep<-cur_rep+1
  
  #first, split trial type in each subject into two random halves
  tmp_dt1<-dt_shr_rt %>% 
    group_by(age_group, id,trialtype) %>%
    mutate(repSet=sample(
      rep(c("A","B"),length.out=n(),size=n())
    ))
  
  #get means for each subject, trialtype, and half
  tmp_dt2<-tmp_dt1 %>% group_by(age_group,id, trialtype, repSet) %>%
    summarise(mean_rt=mean(rt_ms)) #change this into 
  
  #make wider 
  tmp_dt3<-tmp_dt2 %>% 
    pivot_wider(id_cols=c(trialtype, age_group,id),
                names_from=c(repSet),
                values_from=mean_rt)
  
  #get corrs by age group and trial type
  tmp_dt4 <- tmp_dt3 %>% group_by(age_group,trialtype) %>%
    summarise(tmp_meanrt_corr=cor(A,B,use="complete.obs"), # regular corr
              tmp_meanrt_sb=(2*tmp_meanrt_corr)/(1+tmp_meanrt_corr)) #spearman brown 
  
  #add corr and sb to dataframe
  shr_meanrt<-add_row(shr_meanrt,
                    n_rep=seq(cur_rep,cur_rep,length.out=8),
                    age_group=as.factor(tmp_dt4$age_group),
                    trialtype=tmp_dt4$trialtype,
                    meanrt_splithalf=tmp_dt4$tmp_meanrt_corr,
                    meanrt_sb=tmp_dt4$tmp_meanrt_sb)
}

#get final split half estimates 
dtsum_shr_meanrt<-shr_meanrt %>%
  group_by(age_group,trialtype) %>%
  summarise(splithalf=round(mean(meanrt_splithalf),3),
            splithalf5= quantile(meanrt_splithalf,probs=c(.05,.95))[1],
            splithalf95= quantile(meanrt_splithalf,probs=c(.05,.95))[2],
            sb= mean(meanrt_sb),
            sb5= quantile(meanrt_sb,probs=c(.05,.95))[1],
            sb95= quantile(meanrt_sb,probs=c(.05,.95))[2])


#rearrange cols
kable_styling(kable(dtsum_shr_meanrt,digits=4,
                    caption="mean RT split half reliabilities (5000 iterations"))#make table look nice
```

The reliabilities in this example are pretty poor. This is expected, given that the data for each trial was drawn from the same distribution. 

#Error Rates
## Calcuation
error rate = (number of errors + 0.5)/(number of trials + 1) <-from Gonthier et al. 2018

```{r}
#get mean error rate  for each individual and trialtype
dt_err<-dt %>%
  group_by(id,age_group,trialtype) %>%
  summarise(count_acc=sum(acc),
              trial_num=n(),
              count_err=n()-sum(acc)) %>%
  mutate(mean_err=(count_err+0.5)/(trial_num+1))
```

## Variability
```{r}
#view error rates for each trialtype and age group
pirateplot(formula= mean_err~factor(age_group)*factor(trialtype),
           data=dt_err,
           bean.f.o = .3, # customize bean (i.e. violin)
           bean.b.o= 0,
           point.o = 0.3,#customize points
            point.pch = 2,
           point.col = "black",
           point.cex = 0.1,
           inf.f.o = .4, # Inference fill
           #xlab="Age Group",#labels
           ylab="error rates",
           theme=1)
```

## Reliability
```{r rt split half}
#create a df with just the necessary info
dt_shr_err<-dt %>%
  subset(select=c(id,age_group,trialtype,acc)) #might also exclude incorrect trials. Here they've been included in the example. 

#now a df to store output
shr_err<-data.frame(n_rep=integer(),
                    age_group=character(),
                    trialtype=character(),
                       err_splithalf=double(),
                       err_sb=double())

tot_rep<- 5000 #specify number of permutations per subject
cur_rep<-0

while (cur_rep<=tot_rep){
  cur_rep<-cur_rep+1
  
  tmp_dt1<-dt_shr_err %>% 
    group_by(age_group, id,trialtype) %>%
    mutate(repSet=sample(
      rep(c("A","B"),length.out=n(),size=n())
    ))
  
  #get mean RTs for each subject, trialtype, and half
  tmp_dt2<-tmp_dt1 %>% group_by(age_group,id, trialtype, repSet) %>%
    summarise(count_acc=sum(acc),
              trial_num=n(),
              count_err=n()-sum(acc)) %>%
    mutate(mean_err=(count_err+0.5)/(trial_num+1))
  
  #make widerand get corrs
  tmp_dt3<-tmp_dt2 %>% 
    pivot_wider(id_cols=c(trialtype, age_group,id),names_from=c(repSet),
                                   values_from=mean_err)
  
  #get corrs by age group and trial type
  tmp_dt4 <- tmp_dt3 %>% group_by(age_group,trialtype) %>%
    summarise(tmp_err_corr=cor(A,B,use="complete.obs"),
              tmp_err_sb=(2*tmp_err_corr)/(1+tmp_err_corr))
  
  #add corr and sb to dataframe
  shr_err<-add_row(shr_err,
                    n_rep=seq(cur_rep,cur_rep,
                              length.out=nrow(tmp_dt4)),
                    age_group=as.factor(tmp_dt4$age_group),
                    trialtype=tmp_dt4$trialtype,
                    err_splithalf=tmp_dt4$tmp_err_corr,
                    err_sb=tmp_dt4$tmp_err_sb)
}

#get final split half estimates 
dtsum_shr_err<-shr_err %>%
  group_by(age_group,trialtype) %>%
  summarise(splithalf=round(mean(err_splithalf),3),
            splithalf5= quantile(err_splithalf,probs=c(.05,.95))[1],
            splithalf95= quantile(err_splithalf,probs=c(.05,.95))[2],
            sb= mean(err_sb),
            sb5= quantile(err_sb,probs=c(.05,.95))[1],
            sb95= quantile(err_sb,probs=c(.05,.95))[2])


#rearrange cols
kable_styling(kable(dtsum_shr_err,digits=4,
                    caption="AXCPT: mean error rate split half reliabilities (5000 iterations"))#make table look nicet_var="rt_ms")
```

#PBI-RT
## Calcuation
Can swap mean RTs with error rates. 
```{r}
dt_pbirt<-dt %>%
  filter(trialtype=="ay"|trialtype=="bx") %>%
  subset(select=c(id,age_group,trialtype,rt_ms)) %>% 
  group_by(age_group,id,trialtype) %>%
  summarise(meanrt_ms=mean(rt_ms)) %>%
  pivot_wider(id_cols=c(id,age_group),
                names_from=c(trialtype),
                values_from=meanrt_ms) %>%
  mutate(pbirt=(ay - bx)/(ay + bx))
```

## Variability
```{r}
leveneTest(pbirt~as.factor(age_group),data=dt_pbirt,center=mean)

#view a plot
pirateplot(formula= pbirt~factor(age_group),
           data=dt_pbirt,
           bean.f.o = .3, # customize bean (i.e. violin)
           bean.b.o= 0,
           point.o = 0.5,#customize points
            point.pch = 5,
           point.col = "black",
           point.cex = 0.9,
           inf.f.o = .3, # Inference fill
           xlab="Age Group",#labels
           ylab="PBI RT",
           theme=1)

```

## Reliability
```{r}
#create df w/ just necessary info
dt_shr_pbirt<-dt %>%
  filter(trialtype=="ay"|trialtype=="bx") %>%
  subset(select=c(id,age_group,trialtype,rt_ms))

#create an empty dt to hold the results
shr_pbirt<-data.frame(n_rep=integer(),
                       age_group=character(),
                       pbirt_splithalf=double(),
                       pbirt_sb=double())

tot_rep<- 5000 #specify number of repetition
cur_rep<-0 #counter for current repetition

while (cur_rep<=tot_rep){
  cur_rep<-cur_rep+1 #add counter

  #grab only needed trials and split into A and B halves
  tmp_dt1<-dt_shr_pbirt %>% filter(trialtype=="ay"|trialtype=="bx") %>%
    group_by(age_group, id,trialtype) %>%
    mutate(repSet=sample(
      rep(c("A","B"),length.out=n(),size=n())
    ))
  
  #get mean RTs for each subject, trialtype, and half
  tmp_dt2<-tmp_dt1 %>% group_by(age_group,id, trialtype, repSet) %>%
    summarise(meanrt_ms=mean(rt_ms))
  
  #get pbi rt_ms for each half
  tmp_dt3<-tmp_dt2 %>%
    pivot_wider(id_cols=c(id,age_group),
                names_from=c(trialtype, repSet),
                values_from=meanrt_ms) %>%
    mutate(pbirt_A=(ay_A - bx_A)/(ay_A + bx_A),
           pbirt_B=(ay_B - bx_B)/(ay_B + bx_B))
  
  #get corrs by age group
  tmp_dt4 <- tmp_dt3 %>% group_by(age_group) %>%
    summarise(tmp_pbirt_corr=cor(pbirt_A,pbirt_B,use="complete.obs"),
               tmp_pbirt_sb=(2*tmp_pbirt_corr)/(1+tmp_pbirt_corr))
  
  #add corr and sb to dataframe
  shr_pbirt<-add_row(shr_pbirt,
                      n_rep=seq(cur_rep,cur_rep,length.out=2),
                      age_group=as.factor(tmp_dt4$age_group),
                      pbirt_splithalf=tmp_dt4$tmp_pbirt_corr,
                      pbirt_sb=tmp_dt4$tmp_pbirt_sb)
}

#View(shr_pbirt) histograms by age group
ggplot(shr_pbirt, aes(x=pbirt_splithalf,color=age_group))+
  geom_histogram()+
  facet_wrap(~age_group)

#get final split half estimates 
dtsum_shr_pbirt<-shr_pbirt %>%
  group_by(age_group) %>%
  summarise(splithalf=round(mean(pbirt_splithalf),3),
            splithalf5= quantile(pbirt_splithalf,probs=c(.05,.95))[1],
            splithalf95= quantile(pbirt_splithalf,probs=c(.05,.95))[2],
            SpearmanBrown= mean(pbirt_sb),
            SpearmanBrown5= quantile(pbirt_sb,probs=c(.05,.95))[1],
            SpearmanBrown95= quantile(pbirt_sb,probs=c(.05,.95))[2])


#rearrange cols
kable_styling(kable(dtsum_shr_pbirt,digits=4,
                    caption="AXCPT: PBI rt_ms split half reliabilities (5000 iterations"))#make table look nicet_var="rt_ms")
```

#D-context
d'-context= Z(AX hits rate) - Z(BX false alarms rate)= ability to use cue info to drive response

## Calculation
```{r d con}
#get hit rates and FA rates
dt_dcon1<-dt %>% 
  group_by(id,trialtype, age_group) %>% #for each subject and trialtype
  summarise(counts= n(),
            correct=sum(acc)) %>%
  mutate(hit_rate=ifelse(correct/counts==1,.95, #get hit rates and false alarms w/ corrections
                           correct/counts),
         fa_rate=(ifelse((counts-correct)/counts==0,.05,
                  ((counts-correct)/counts))))

#now keep going
dt_dcon2<-dt_dcon1 %>%
  filter(trialtype=="ax"|trialtype=="bx") %>% #filter out AY and BY trials
  pivot_wider(id_cols = c(id,age_group), 
              names_from = trialtype, 
              values_from = c(hit_rate,fa_rate)) %>%
  mutate(dcon = (qnorm(hit_rate_ax) - qnorm(fa_rate_bx))) #%>%

# final dt_dcon
dt_dcon<-dt_dcon2

#view dcontext distribution
hist(dt_dcon$dcon,breaks=20)

```

## Variability
```{r dcontvariability}
leveneTest(dcon~as.factor(age_group),data=dt_dcon)

#create a graph of observed variances across trialtypes by age groups
pirateplot(formula= dcon~factor(age_group),
           data=dtsum,
           bean.f.o = .3, # customize bean (i.e. violin)
           bean.b.o= 0,
           point.o = 0.9,#customize points
            point.pch = 21,
           point.col = "black",
           point.cex = 1,
           inf.f.o = .3, # Inference fill
           xlab="Age Group",#labels
           ylab="d-context",
           theme=1)

```

##Reliability
```{r dcon rel}
#create df w/ just necessary info 
dt_shr_dcon<-dt %>%
  filter(trialtype=="ax"|trialtype=="bx") %>%
  subset(select=c(id,age_group,trialtype,acc))

#empty df to store results
shr_dcon<-data.frame(n_rep=integer(),
                       age_group=character(),
                       dcon_splithalf=double(),
                       dcon_sb=double())

tot_rep<- 5000 #specify number of permutations 
cur_rep<-0

while (cur_rep<=tot_rep){
  cur_rep<-cur_rep+1

  tmp_dt1<-dt_shr_dcon %>%
    group_by(age_group, id,trialtype) %>%
    mutate(repSet=sample(
      rep(c("A","B"),length.out=n(),size=n())
    ))
  
  #get mean RTs for each subject, trialtype, and half
  tmp_dt2<- tmp_dt1 %>% 
    group_by(age_group,trialtype,id,repSet) %>%
    summarise(counts= n(),
            correct=sum(acc)) %>%
    mutate(hit_rate=ifelse(correct/counts==1,.95,
                           correct/counts),
         fa_rate=(ifelse((counts-correct)/counts==0,.05,
                  ((counts-correct)/counts)))) #get hit rates and false alarms

  #now keep going
  tmp_dt3<-tmp_dt2 %>%
    pivot_wider(id_cols = c(id,repSet,age_group), 
                names_from = c(trialtype), 
                values_from = c(hit_rate,fa_rate)) %>%
    mutate(dcon = (qnorm(hit_rate_ax) - qnorm(fa_rate_bx))) %>%
    pivot_wider(id_cols=c(id,age_group),
                names_from=repSet,
                values_from=dcon)
  
  #get corrs by age group
  tmp_dt4 <- tmp_dt3 %>% group_by(age_group) %>%
    summarise(tmp_dcon_corr=cor(A,B,use="complete.obs"),
               tmp_dcon_sb=(2*tmp_dcon_corr)/(1+tmp_dcon_corr))
  
  #add corr and sb to dataframe
  shr_dcon<-add_row(shr_dcon,
                      n_rep=c(cur_rep, cur_rep),
                      age_group=as.factor(tmp_dt4$age_group),
                      dcon_splithalf=tmp_dt4$tmp_dcon_corr,
                      dcon_sb=tmp_dt4$tmp_dcon_sb)
}

#View(shr_dcon) histograms by age group
ggplot(shr_dcon, aes(x=dcon_splithalf,color=age_group))+
  geom_histogram()+
  facet_wrap(~age_group)

#save output 
# write.csv(shr_dcon,"05_Psychometrics/SearchCosts_AXCPT_dconSHR_201019.csv",row.names=FALSE)

#get final split half estimates 
dtsum_shr_dcon<-shr_dcon %>%
  group_by(age_group) %>%
  summarise(splithalf=round(mean(dcon_splithalf),3),
            splithalf5= quantile(dcon_splithalf,probs=c(.05,.95))[1],
            splithalf95= quantile(dcon_splithalf,probs=c(.05,.95))[2],
            sb= mean(dcon_sb),
            sb5= quantile(dcon_sb,probs=c(.05,.95))[1],
            sb95= quantile(dcon_sb,probs=c(.05,.95))[2])

#rearrange cols
kable_styling(kable(dtsum_shr_dcon,digits=4,
                    caption="AXCPT: dcontext split half reliabilities (5000 iterations"))#make table look nicet_var="logRT")
```

# A-cue bias
A-cue bias= 1/2*(Z(AX hits rate)+Z(AY false alarms rate))= tendency to make target response to A cues

##Calculation
```{r}
#A-cue bias= 1/2*(Z(AX hits rate)+Z(AY false alarms rate))= tendency to make target response to A cues
#get hit rates and FA rates
dt_abias1<-dt %>% 
  group_by(id,trialtype,age_group) %>% #for each subject and trialtype
  summarise(counts= n(),
            correct=sum(acc)) %>%
  mutate(hit_rate=ifelse(correct/counts==1,.95, #get hit rates and false alarms w/ corrections
                           correct/counts),
         fa_rate=(ifelse((counts-correct)/counts==0,.05,
                  ((counts-correct)/counts))))

#now keep going
dt_abias2<-dt_abias1 %>%
  filter(trialtype=="ax"|trialtype=="ay") %>% #filter out BX and BY trials
  pivot_wider(id_cols = c(id,age_group), 
              names_from = trialtype, 
              values_from = c(hit_rate,fa_rate)) %>%
  mutate(abias = 0.5*(qnorm(hit_rate_ax) - qnorm(fa_rate_ay))) #%>%

# final dt_dcon
dt_abias<-dt_abias2

#view dcontext distribution
hist(dt_abias$abias,breaks=20)
```

## Variability
```{r}
leveneTest(abias~as.factor(age_group),data=dt_abias)


#create a graph of observed variances across trialtypes by age groups
pirateplot(formula= abias~factor(age_group),
           data=dt_abias,
           bean.f.o = .3, # customize bean (i.e. violin)
           bean.b.o= 0,
           point.o = 0.9,#customize points
            point.pch = 21,
           point.col = "black",
           point.cex = 1,
           inf.f.o = .3, # Inference fill
           xlab="Age Group",#labels
           ylab="A-cue bias",
           theme=1)
```

## Reliability
```{r A cue bias rel}
#A-cue bias= 1/2*(Z(AX hits rate)+Z(AY false alarms rate))= tendency to make target response to A cues
#create df w/ just necessary info 
dt_shr_abias<-dt %>%
  filter(trialtype=="ax"|trialtype=="ay") %>%
  subset(select=c(id,age_group,trialtype,acc))

shr_abias<-data.frame(n_rep=integer(),
                       age_group=character(),
                       abias_splithalf=double(),
                       abias_sb=double())

tot_rep<- 5000 #specify number of permutations 
cur_rep<-0

while (cur_rep<=tot_rep){
  cur_rep<-cur_rep+1

  tmp_dt1<-dt_shr_abias %>%
    group_by(age_group, id,trialtype) %>%
    mutate(repSet=sample(
      rep(c("A","B"),length.out=n(),size=n())
    ))
  
  #get mean RTs for each subject, trialtype, and half
  tmp_dt2<- tmp_dt1 %>% 
    group_by(age_group,trialtype,id,repSet) %>%
    summarise(counts= n(),
            correct=sum(acc)) %>%
    mutate(hit_rate=ifelse(correct/counts==1,.95, #get hit rates and false alarms w/ corrections
                           correct/counts),
         fa_rate=(ifelse((counts-correct)/counts==0,.05,
                  ((counts-correct)/counts))))

  #now keep going and do calcs
  tmp_dt3<-tmp_dt2 %>%
    pivot_wider(id_cols = c(id,repSet,age_group), 
                names_from = c(trialtype), 
                values_from = c(hit_rate,fa_rate)) %>%
    mutate(abias = 0.5*(qnorm(hit_rate_ax) - qnorm(fa_rate_ay))) %>%
    pivot_wider(id_cols=c(id,age_group),
                names_from=repSet,
                values_from=abias)
  
  #get corrs by age group
  tmp_dt4 <- tmp_dt3 %>% group_by(age_group) %>%
    filter(is.finite(A) & is.finite(B)) %>%
    summarise(tmp_abias_corr=cor(A,B,use="complete.obs"),
               tmp_abias_sb=(2*tmp_abias_corr)/(1+tmp_abias_corr))
  
  #add corr and sb to dataframe
  shr_abias<-add_row(shr_abias,
                      n_rep=c(cur_rep, cur_rep),
                      age_group=as.factor(tmp_dt4$age_group),
                      abias_splithalf=tmp_dt4$tmp_abias_corr,
                      abias_sb=tmp_dt4$tmp_abias_sb)
}

#View(shr_abias) histograms by age group
ggplot(shr_abias, aes(x=abias_splithalf,color=age_group))+
  geom_histogram()+
  facet_wrap(~age_group)

#get final split half estimates 
dtsum_shr_abias<-shr_abias %>%
  group_by(age_group) %>%
  summarise(splithalf=round(mean(abias_splithalf),3),
            splithalf5= quantile(abias_splithalf,probs=c(.05,.95))[1],
            splithalf95= quantile(abias_splithalf,probs=c(.05,.95))[2],
            sb= mean(abias_sb),
            sb5= quantile(abias_sb,probs=c(.05,.95))[1],
            sb95= quantile(abias_sb,probs=c(.05,.95))[2])

#rearrange cols
kable_styling(kable(dtsum_shr_abias,digits=4,
                    caption="AXCPT: A-cue bias split half reliabilities (5000 iterations"))#make table look nicet_var="logRT")
```